Semi-quantitative petrographic characterization of Bronze Age quartzite tools – a multivariate approach

### 2021-09-10 analysis/paper/paper.Rmd
added check.names = FALSE which removes need for substringing data when transposing it. Working on adding spectra plots for raman and nir

#Import xrf data
XRF.csv <-
  read.csv2("./analysis/data/raw_data/XRF/XRF_quantitative.csv", sep = ";", dec = ".", header = TRUE, na = c("","NA","NULL"))

#Import nir data, set empty fields to NA
nir.csv <-
  read.csv2("./analysis/data/derived_data/NIR_master_averaged_20210827.csv", sep = ";",
            dec = ",", header = TRUE, na = c("","NA","NULL",NULL))

#Import raman data, set empty fields to NA
raman.csv <-
  read.csv2("./analysis/data/derived_data/raman_baseline_corrected_20210826.csv", sep = ";", dec = ",", header = TRUE, na = c("","NA","NULL",NULL))

#Import sample list, set empty fields to NA
samples <-
  read.csv2("./analysis/data/raw_data/bifacial_points.csv", sep = ";", dec = ".", header = TRUE, na.strings=c("", "NA"))

@@

The Raman instrumentation used was a portable i-Raman EX featuring a 1064 nm excitation laser. The spectral coverage of the instrumentation is 175 – 2 500 cm-1 ^[Instead of reporting Raman measurements in spectral wavelength (nm) they are typically given in wavenumber (cm-1), referred to as *Raman shift*] with a resolution of 9.5 cm-1. Access to the spectrometer was made possible by the department of Forest Biomaterials and Technology at the Swedish University of Agricultural Sciences (SLU) in Umeå.

@@

@@

```{r xrf-violin-plot, fig.width=15, fig.height=15, fig.fullwidth=TRUE, fig.cap="Violin plot of element content in quartzite samples measured using pXRF. **n** represents no. of measurements above Limit of Detection (LOD) for that element out of a total of 478 measurements. Note the bimodal distribution of Si measurements, these groups are reflected in the score plot of the PCA"}
#Prepare data in long format for violin plots
Points_long <-
  Points.xrf %>%
  select(X12Mg,X13Al, X14Si, X15P, X16S, X19K, X20Ca, X23V, X25Mn, X26Fe, X30Zn, X38Sr, X39Y, X40Zr, X56Ba, X81Ti) %>%
  gather(key = "Element", value = "Value")

#Remove initial X introduced by read.csv to column names
Points_long$Element <- substring(Points_long$Element, 2)

@@

```{r xrf-loadings, fig.width=8, fig.height=8, fig.cap="The three first loading vectors (P1-P3) with scores for each loading (element)"}
#select elements to include in pca, as well as columns to group by
xrf <-
  Points.xrf %>%
    select(reading_no, sample_id, munsell_hue, material, X12Mg, X13Al, X14Si, X15P, X16S, X19K, X20Ca, X23V, X25Mn, X26Fe, X30Zn, X38Sr, X39Y, X40Zr, X56Ba, X81Ti)

@@

#Extract loadings from xrf.pca
loadings <- as.data.frame(xrf.pca$rotation[,1:3])
#As elements are stored as rownames, set them as first column so it can be used as a variable
loadings <- setDT(loadings, keep.rownames = "Element")[]
# remove inital 'X' from element names, which was introduced in initial import to R
loadings$Element <- substring(loadings$Element, 2)
#Melt loadings from wide to long format
loadings.melted <- melt(loadings, id.vars="Element")
#Set names of columns
loadings <- setNames(loadings.melted, c("Element", "PC", "Value"))

>>>>

#Import xrf data
XRF.csv <-
  read.csv2("./analysis/data/raw_data/XRF/XRF_quantitative.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na = c("","NA","NULL"))

#Import nir data, set empty fields to NA
nir.csv <-
  read.csv2("./analysis/data/derived_data/NIR_master_averaged_20210827.csv", sep = ";",
            dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))

#Import raman data, set empty fields to NA
raman.csv <-
  read.csv2("./analysis/data/derived_data/raman_baseline_corrected_20210826.csv", sep = ";", dec = ",", header = TRUE, check.names = FALSE, na = c("","NA","NULL",NULL))

#Import sample list, set empty fields to NA
samples <-
  read.csv2("./analysis/data/raw_data/bifacial_points.csv", sep = ";", dec = ".", header = TRUE, check.names = FALSE, na.strings=c("", "NA"))

@@

The Raman instrumentation used was a portable i-Raman EX featuring a 1064 nm excitation laser. The spectral coverage of the instrumentation is 175 – 2 500 cm-1 ^[Raman shifts are typically given in wavenumber (cm-1), instead of spectral wavelength (nm)] with a resolution of 9.5 cm-1. Access to the spectrometer was made possible by the department of Forest Biomaterials and Technology at the Swedish University of Agricultural Sciences (SLU) in Umeå.

@@

```{r nir-spectra, fig.width=20, fig.height=8, fig.cap= "Average NIR spectra by hue. Darker samples feature higher overall absorbance, as well as less pronounced peaks, when compared to lighter/translucent samples."}

#Aggregate raw NIR data by inferred hue
nir.agg <- 
  aggregate(nir.csv[, 28:2178], list(nir.csv$hue), mean)

#Transpose the data in preparation for working in ChemoSpec 
# make row names unique & sorted correctly in increasing order
# by appending numbers in running order
nir.t <- nir.agg %>%
  dplyr::mutate(Group.1 = paste(seq(1, n()),
                         Group.1 ,
                      sep = "_")) %>%
  gather(row, value, -Group.1 ) %>% 
  spread(Group.1 , value)

# strip away the appended numbers from the newly created column names
names(nir.t) <- sapply(strsplit(names(nir.t), "_"), function(x){x[2]})
#Change first column name to 'freq'  
colnames(nir.t)[1] <- "freq"

#Set the first column containing frequencies to numeric
nir.t[,1] <- sapply(nir.t[,1], as.numeric)
#Sort dataframe by frequencies asc.
nir.t <- nir.t[order(nir.t$freq),] 
#Write to csv file, since ChemoSpec doesn't work with dataframes
write.csv2(nir.t,"./analysis/data/derived_data/nir_averaged_by_hue.csv", row.names = FALSE)

#Read exported csv and create new file for chemospec
nir <-matrix2SpectraObject(
  gr.crit = c("Translucent", "White", "Dark", "Light" ),
  gr.cols = c("purple", "blue", "black", "red"),
  freq.unit ="Wavelength (nm)",
  int.unit ="Absorbance",
  descrip ="Bifacial points measurements",
  in.file = "./analysis/data/derived_data/NIR_averaged_by_hue.csv",
  sep = ";",
  dec = ",",
  chk = TRUE,
  out.file = "nir")

#Plot spectra averaged by inferred hue
Dark <- plotSpectra(nir,
            which = 1,
            yrange = c(1.05,1.2),
            lab.pos = 0.0,
            main = "Average spectra - Dark", font.lab = 2)
#Plot spectra averaged by inferred hue
Light <- plotSpectra(nir,
            which = 2,
            yrange = c(0.7,1),
            lab.pos = 0.0,
            main = "Average spectra - Dark", font.lab = 2)
#Plot spectra averaged by inferred hue
Translucent <- plotSpectra(nir,
            which = 3,
            yrange = c(1.05,1.2),
            lab.pos = 0.0,
            main = "Average spectra - Dark", font.lab = 2)
#Plot spectra averaged by inferred hue
White <- plotSpectra(nir,
            which = 4,
            yrange = c(1.05,1.2),
            lab.pos = 0.0,
            main = "Average spectra - Dark", font.lab = 2)
grid.arrange()
```

@@

```{r xrf-violin-plot, fig.width=15, fig.height=15, fig.fullwidth=TRUE, fig.cap="Violin plot of element content in quartzite samples measured using pXRF. **n** represents no. of measurements above Limit of Detection (LOD) for that element out of a total of 478 measurements. Note the bimodal distribution of Si measurements, these groups are reflected in the score plot of the PCA"}
#Prepare data in long format for violin plots
Points_long <-
  Points.xrf %>%
  select(`12Mg`,`13Al`, `14Si`, `15P`, `16S`, `19K`, `20Ca`, `23V`, `25Mn`, `26Fe`, `30Zn`, `38Sr`, `39Y`, `40Zr`, `56Ba`, `81Ti`) %>%
  gather(key = "Element", value = "Value")

@@

```{r xrf-loadings, fig.width=8, fig.height=8, fig.cap="The three first loading vectors (P1-P3) with scores for each loading (element)"}
#select elements to include in pca, as well as columns to group by
xrf <-
  Points.xrf %>%
    select(reading_no, sample_id, munsell_hue, material, `12Mg`, `13Al`, `14Si`, `15P`, `16S`, `19K`, `20Ca`, `23V`, `25Mn`, `26Fe`, `30Zn`, `38Sr`, `39Y`, `40Zr`, `56Ba`, `81Ti`)

@@

#Extract loadings from xrf.pca
loadings <- as.data.frame(xrf.pca$rotation[,1:3])
#As elements are stored as rownames, set them as first column so it can be used as a variable
loadings <- setDT(loadings, keep.rownames = "Element")[]
#Melt loadings from wide to long format
loadings.melted <- melt(loadings, id.vars="Element")
#Set names of columns
loadings <- setNames(loadings.melted, c("Element", "PC", "Value"))
